---
title: "Final 222 Project"
output:
  pdf_document: default
  html_document: default
date: "2026-03-02"
---

Libraries

```{r}
library(nhanesA)
library(purrr)
library(dplyr)
library(ggplot2)
library(car)
library(survey)
library(svydiags)
```

Retrieving Nhanes Data

```{r}
dfs = list(
  demographics = nhanes("DEMO_I"),
  chloestral   = nhanes("BPQ_I"),
  cardio       = nhanes("CDQ_I"),
  diabetes     = nhanes("DIQ_I"),
  diet         = nhanes("DBQ_I"),
  disability   = nhanes("DLQ_I"),
  drugs        = nhanes("DUQ_I"),
  food_sec     = nhanes("FSQ_I"),
  housing      = nhanes("HOQ_I"),
  income       = nhanes("INQ_I"),
  conditions   = nhanes("MCQ_I"),
  activity     = nhanes("PAQ_I"),
  kidney       = nhanes("KIQ_U_I"),
  occuptation  = nhanes("OCQ_I"),
  mental       = nhanes("DPQ_I")
)

df = reduce(dfs, left_join, by = "SEQN")
```

Creating Dataframe

```{r}
str(df)
## creating new cardiovascular health outcome
df <- df %>%
  mutate(
    cardio = if_else(
      MCQ160B == "Yes" |
      MCQ160C == "Yes" |
      MCQ160D == "Yes" |
      MCQ160E == "Yes",
      1, 0
    )
  )

## only selecting columns with variables we are going to put into the model
# SDMVPSU: Primary Sampling Unit
# SDMVSTRA: Sampling Stratum
# WTINT2YR: Interview Weight For 2-Year NHANES Cycle
df <- df %>%
  dplyr::select(SDMVPSU,SDMVSTRA, WTINT2YR, RIDRETH3, RIDAGEYR, RIAGENDR, INDFMPIR, DMDHREDU, cardio)

## getting rid of any rows with NA values
df <- na.omit(df)

## selecting rows only if they are at least 20 years old (since cardiovascular variables were only for people above 20)
df <- df %>%
  filter(RIDAGEYR >= 20)

## rename columns to make more sense 
df <- df %>%
  rename(
    race = RIDRETH3,
    age = RIDAGEYR,
    gender = RIAGENDR,
    income = INDFMPIR,
    education = DMDHREDU
  )

## combining mexican american and other hispanic into just one category: hispanic
df <- df %>%
  mutate(
    race = case_when(
      race %in% c("Mexican American", "Other Hispanic") ~ "Hispanic",
      TRUE ~ race
    )
  )

## getting rid of rows with education as don't know
df <- df %>%
  filter(education != "Don't Know")

## combining 9-11th grade and less than 9th grade 
df <- df %>%
  mutate(
    education = case_when(
      education %in% c(
        "Less Than 9th Grade",
        "9-11th Grade (Includes 12th grade with no diploma)"
      ) ~ "Not High School Graduate",
      TRUE ~ education
    )
  )

## combining Some College or AA degree and High School Grad/GED or Equivalent
df <- df %>%
  mutate(
    education = case_when(
      education %in% c("High School Grad/GED or Equivalent",
                       "Some College or AA degree") ~ "High School Graduate",
      TRUE ~ education
    )
  )

```

Analysis

```{r}
## checking variable type 
df$cardio    <- as.factor(df$cardio)
df$education <- as.factor(df$education)
df$gender    <- as.factor(df$gender)
df$race      <- as.factor(df$race)

## changing reference variables 
df$education <- relevel(as.factor(df$education),
                        ref = "Not High School Graduate")

df$race <- relevel(as.factor(df$race),
                   ref = "Non-Hispanic White")
```

```{r}
## exploratory data analysis 

df %>%
  group_by(education) %>%
  summarise(prop_cardio = mean(cardio == 1)) %>%
  ggplot(aes(x = education, y = prop_cardio)) +
  geom_col() +
  labs(
    x = "Education Level",
    y = "Proportion with Cardiovascular Disease",
    title = "Cardiovascular Disease by Education Level"
  ) +
  theme_minimal()
```

```{r}
## fitted the model with just the confounders, all have p-values less than 0.2 besides  raceNon-Hispanic Black (p = 0.50062) and raceOther Race - Including Multi-Racial (p = 0.32112), so I'm going to keep all confounder 

model_confounders <- glm(
  cardio ~ age + gender + race + income,
  data = df,
  family = binomial
)
summary(model_confounders)
```

```{r}
model_base <- glm(
  cardio ~ education + age + gender + race + income,
  data = df,
  family = binomial
)
summary(model_base)
```

```{r}
## cardiovascular outcomes based on education really don't change by race 
model_int_race <- glm(
  cardio ~ education * race + age + gender + income,
  data = df,
  family = binomial
)

summary(model_int_race)
```

```{r}
## cardiovascular outcomes based on education do change by gender
model_int_gender <- glm(
  cardio ~ education * gender + age + race + income,
  data = df,
  family = binomial
)

summary(model_int_gender)
```

```{r}
anova(model_base, model_int_gender, test = "LRT")

```

*investigating whether the effect of education differs by gender adjusted for age, race, income*

*education reference = not high school graduate, gender reference = male*

**Among males, there is no statistically significant association between education level and cardiovascular disease**

-   educationCollege Graduate or above 0.0878 p = 0.7017

-   educationHigh School Graduate 0.1549 p = 0.3881

**The education effect for females differs from males by -0.9027 (p = 0.0117 ) (on the log-odds scale)**

-   educationCollege Graduate or above:genderFemale (Estimate = -0.9027, p = 0.0117)

-   female college effect = exp(0.087806 + (-0.9027)) = exp(-0.814894) = 0.4426862

**Among females, college graduates have approximately 56% lower odds of cardiovascular disease compared to women without a high school degree.**

**There is statistical evidence that the association between education and cardiovascular health differs by gender.**

```{r}
#Binning Continuous Explanatory Variables 
df = df%>%
  mutate(
    age_bin = case_when(
      age >= 20  & age < 40 ~"20-39",
      age >= 40 & age < 60 ~"40-59",
      age >= 60 ~ "60+",
    ),
    income_bin = case_when(
      income < 1 ~"Low",
      income >= 1 & income <= 4 ~"Middle",
      income > 4 ~ "High"
    )
  )
df$age_bin = factor(df$age_bin, levels= c("20-39","40-59","60+"))
df$income_bin = factor(df$income_bin, levels = c("Low","Middle","High"))
```

GLM Model Diagnostics 
```{r}
#Aggregating Data 
df$cardio_num = ifelse(df$cardio == 1|df$cardio == "1",1,0)
ag.df = aggregate(cardio_num~education+gender+race+age_bin+income_bin,
                  data = df,
                  FUN = sum)
ag.df$tot = aggregate(cardio_num~education+gender+race+age_bin+income_bin,
                  data = df,
                  FUN = length)$cardio_num
head(ag.df)
```


```{r}
#Aggregate Data Logistic Regression
ag_mod <- glm(cardio_num/tot~education + age_bin + gender + race + income_bin,
             family = binomial,
             data = ag.df,weights = tot)

summary(ag_mod)
```

```{r}
ag_mod_race <- glm(cardio_num/tot~education*race+gender+age_bin+income_bin,
             family = binomial,
             data = ag.df,weights = tot)

summary(ag_mod_race)
```

```{r}
ag_mod_gender = glm(cardio_num/tot~education*gender+race+age_bin+income_bin,
             family = binomial,
             data = ag.df,weights = tot)

summary(ag_mod_gender)
```

```{r}
anova(ag_mod, ag_mod_gender, test = "LRT")
```

```{r}
###There is no evidence of lack of fit (0.91>0.05)
dev = deviance(ag_mod_gender)
res = df.residual(ag_mod_gender)
pchisq(dev,res,lower.tail = FALSE)
```

```{r}
###There is no evidence of lack of fit (0.093>0.05)
pearson_chisq = sum(residuals(ag_mod_gender,type="pearson")^2)
pchisq(pearson_chisq, res, lower.tail = FALSE)
```

```{r}
###Standard GLM Diagnostic Plots to look for unusual residual patterns, 
###high leverage points, and possible influential observations

par(mfrow = c(2,2))
plot(ag_mod_gender)
par(mfrow=c(1,1))
```
```{r}
#Influence Diagnostics
Delta.Chi = rstudent(ag_mod_gender, type = "pearson")
CD = cooks.distance(ag_mod_gender)
iMI = influence.measures(ag_mod_gender)
fit = ag_mod_gender$fitted.values
```

```{r}
h = hatvalues(ag_mod_gender)
plot(h,Delta.Chi,
     main = "Residual vs Leverage",
     xlab = "Leverage",
     ylab = "Studentized Pearson Residuals")
```

```{r}
plot(fit,CD,
     main = "Cooks Distance vs Fitted Probabilities",
     xlab = "Fitted Probabilities",
     ylab = "Cooks Distance")
```

```{r}
plot(fit,Delta.Chi^2,
     main = "Squared Studentized Pearson vs Fitted Probabilities",
     xlab = "Fitted Probabilities",
     ylab = "Squared Studentized Pearson")
```
**the residual and influence plots were used to check for unusual patterns, outliers, and observations with high leverage overall, there was no major pattern suggesting serious model misspecification**
```{r}
influencePlot(ag_mod_gender)
```

```{r}
#Sensitivity Check
#Removing observation 62 does not alter the conclusion(B = -0.9376, p = 0.0083)
m1 = glm(cardio_num/tot~education*gender+race+age_bin+income_bin,
         data = ag.df[-62,],
         family = binomial,weights = tot)
round(summary(m1)$coefficients,4)
```

```{r}
#Removing observation 146 does not alter the conclusion(B = -0.9417, p = 0.0079)
m2 = glm(cardio_num/tot~education*gender+race+age_bin+income_bin,
         data = ag.df[-146,],
         family = binomial,weights = tot)
round(summary(m2)$coefficients,4)
```

```{r}
#Removing observation 148 does not alter the conclusion(B = -1.1675, p = 0.0029)
m3 = glm(cardio_num/tot~education*gender+race+age_bin+income_bin,
         data = ag.df[-148,],
         family = binomial,weights = tot)
round(summary(m3)$coefficients,4)
```

```{r}
#Removing observation 149 does not alter the conclusion(B = -0.9416 , p = 0.0080)

m4 = glm(cardio_num/tot~education*gender+race+age_bin+income_bin,
         data = ag.df[-149,],
         family = binomial,weights = tot)
round(summary(m4)$coefficients,4)
```

```{r}
#Removing observation 164 does not alter the conclusion(B = -0.9499 , p = 0.0074)
#raceNon-Hispanic Black (p = 0.120) becomes NON-SIGNIFICANT
#The Non-Hispanic Black race finding is sensitive to this one observation
m5 = glm(cardio_num/tot~education*gender+race+age_bin+income_bin,
         data = ag.df[-164,],
         family = binomial,weights = tot)
round(summary(m5)$coefficients,4)
```

```{r}
#Removing observation 252 does not alter the conclusion(B = -0.9394 , p = 0.0081)

m6 = glm(cardio_num/tot~education*gender+race+age_bin+income_bin,
         data = ag.df[-252,],
         family = binomial,weights = tot)
round(summary(m6)$coefficients,4)
```
*Overall the model appears to fit the data reasonably well,and the main interaction result for college education by gender is stable*

Survey Data Analysis

```{r}
#Collapsed race for survey interaction models b/c survey-weighted interaction models with 5 race categories use too many parameters relative to the available survey design df

df <- df%>%
  mutate(race_collapsed = case_when(
    race=="Non-Hispanic White"~"Non-Hispanic White",
    TRUE~"Minority"
  ))
df$race_collapsed = relevel(as.factor(df$race_collapsed),
                  ref = "Non-Hispanic White" )

nhc = svydesign(
  id =~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTINT2YR,
  data = df,
  nest = TRUE
)
```

```{r}
# Used Quasibinomial because survey weights cause the standard binomial variance to be misspecified 
### All have p-values less than 0.2 besides race_collapsedMinority (p=0.77)
model_confounders_svy <- svyglm(
  cardio ~ age + gender + race_collapsed + income,
  design = nhc,
  family = quasibinomial()
)
summary(model_confounders_svy)

model_confounders_svy <- svyglm(
  cardio ~ age + gender + race + income,
  design = nhc,
  family = quasibinomial()
)
summary(model_confounders_svy)
```

```{r}
model_base_svy = svyglm(cardio~education+age+gender+race+income, 
                        design = nhc,
                        family = quasibinomial())
summary(model_base_svy)
```

```{r}
##P-value is NaN 
model_int_race_svy <- svyglm(
  cardio ~ education * race + age + gender + income,
  design = nhc,
  family = quasibinomial()
)

summary(model_int_race_svy)
```

```{r}
## cardiovascular outcomes based on education really don't change by race(p-value = 1.63)

model_int_race_svy <- svyglm(
  cardio ~ education * race_collapsed + age + gender + income,
  design = nhc,
  family = quasibinomial()
)

summary(model_int_race_svy)
```

```{r}
## cardiovascular outcomes based on education change by gender

model_int_gender_svy <- svyglm(
  cardio ~ education * gender + age + race_collapsed + income,
  design = nhc,
  family = quasibinomial()
)

summary(model_int_gender_svy)
```

```{r}
regTermTest(model_int_race_svy, ~education:race_collapsed)
```

```{r}
regTermTest(model_int_gender_svy, ~education:gender)
```
*investigating whether the effect of education differs by gender adjusted for age, race, income using survey-weighted analysis*

*education reference = not high school graduate, gender reference = male*

**Among males, there is no statistically significant association between education level and cardiovascular disease**

-   educationCollege Graduate or above -0.2098, p = 0.5114

-   educationHigh School Graduate -0.0430, p = 0.8364

**The education effect for females differs from males by -1.0127 (p = 0.06601 ) (on the log-odds scale)**

-   educationCollege Graduate or above:genderFemale (Estimate = -1.0127, p = 0.06601)

-   female college effect = exp(-0.209773 + (-1.012664)) = exp(-1.222437) = 0.2945

**Among females, college graduates have approximately 71% lower odds of cardiovascular disease compared to women without a high school degree.**

**There is weak statistical evidence that the association between education and cardiovascular health differs by gender after accounting for the survey design.**


Survey Model Diagnostics
```{r}
#stratified clustered design
#Few sampled clusters may have a noticeable influnence on the fitted model
cd = svyCooksD(model_int_gender_svy,"SDMVSTRA","SDMVPSU",doplot = TRUE)
summary(cd)
```

```{r}
#No widespread instability across model coefficients 
dfb = svydfbetas(model_int_gender_svy,"SDMVSTRA","SDMVPSU",3)
apply(abs(dfb$Dfbetas)>dfb$cutoff,1,sum)
```

```{r}
#Model has some sensitivity to influential observations
dff=svydffits(model_int_gender_svy,"SDMVSTRA","SDMVPSU",4)
sum(abs(dff$Dffits) > dff$cutoff)
```

```{r}
#Several Observations are above the red cutoff line
#May mean that some observations have more influence on the model
h = svyhat(model_int_gender_svy,doplot = TRUE)
```

```{r}
#Some large standardized residuals
res = svystdres(model_int_gender_svy, "SDMVSTRA","SDMVPSU", doplot = TRUE)
```

```{r}
#No strong evidence of severe multicollinearity
X = model.matrix(~education*gender+age+race_collapsed+income,
                 data = model.frame(model_int_gender_svy))
rows_used = as.numeric(names(model_int_gender_svy$y))
w = weights(nhc)[rows_used]

svyvif(model_int_gender_svy, X[,-1], w,"SDMVSTRA","SDMVPSU")
```

```{r}
#No strong evidence of severe collinearity
svycollinear(model_int_gender_svy,X,w,sc=TRUE, rnd=3, fuzz=0.05)
```
*the survey-weighted education by gender model shows some influential observations, some higher-leverage points, and some large residuals,but there is no clear evidence that the model is invalid.*