---
title: "Final 222 Project"
output:
  pdf_document: default
  html_document: default
date: "2026-03-02"
---

Libraries

```{r}
library(nhanesA)
library(purrr)
library(dplyr)
library(ggplot2)
```

Retrieving Nhanes Data

```{r}
dfs = list(
  demographics = nhanes("DEMO_I"),
  chloestral   = nhanes("BPQ_I"),
  cardio       = nhanes("CDQ_I"),
  diabetes     = nhanes("DIQ_I"),
  diet         = nhanes("DBQ_I"),
  disability   = nhanes("DLQ_I"),
  drugs        = nhanes("DUQ_I"),
  food_sec     = nhanes("FSQ_I"),
  housing      = nhanes("HOQ_I"),
  income       = nhanes("INQ_I"),
  conditions   = nhanes("MCQ_I"),
  activity     = nhanes("PAQ_I"),
  kidney       = nhanes("KIQ_U_I"),
  occuptation  = nhanes("OCQ_I"),
  mental       = nhanes("DPQ_I")
)

df = reduce(dfs, left_join, by = "SEQN")
```

Creating Dataframe

```{r}
## creating new cardiovascular health outcome
df <- df %>%
  mutate(
    cardio = if_else(
      MCQ160B == "Yes" |
      MCQ160C == "Yes" |
      MCQ160D == "Yes" |
      MCQ160E == "Yes",
      1, 0
    )
  )

## only selecting columns with variables we are going to put into the model
df <- df %>%
  select(RIDRETH3, RIDAGEYR, RIAGENDR, INDFMPIR, DMDHREDU, cardio)

## getting rid of any rows with NA values
df <- na.omit(df)

## selecting rows only if they are at least 20 years old (since cardiovascular variables were only for people above 20)
df <- df %>%
  filter(RIDAGEYR >= 20)

## rename columns to make more sense 
df <- df %>%
  rename(
    race = RIDRETH3,
    age = RIDAGEYR,
    gender = RIAGENDR,
    income = INDFMPIR,
    education = DMDHREDU
  )

## combining mexican american and other hispanic into just one category: hispanic
df <- df %>%
  mutate(
    race = case_when(
      race %in% c("Mexican American", "Other Hispanic") ~ "Hispanic",
      TRUE ~ race
    )
  )

## getting rid of rows with education as don't know
df <- df %>%
  filter(education != "Don't Know")

## combining 9-11th grade and less than 9th grade 
df <- df %>%
  mutate(
    education = case_when(
      education %in% c(
        "Less Than 9th Grade",
        "9-11th Grade (Includes 12th grade with no diploma)"
      ) ~ "Not High School Graduate",
      TRUE ~ education
    )
  )

## combining Some College or AA degree and High School Grad/GED or Equivalent
df <- df %>%
  mutate(
    education = case_when(
      education %in% c("High School Grad/GED or Equivalent",
                       "Some College or AA degree") ~ "High School Graduate",
      TRUE ~ education
    )
  )

```

Analysis

```{r}
## checking variable type 
df$cardio    <- as.factor(df$cardio)
df$education <- as.factor(df$education)
df$gender    <- as.factor(df$gender)
df$race      <- as.factor(df$race)

## changing reference variables 
df$education <- relevel(as.factor(df$education),
                        ref = "Not High School Graduate")

df$race <- relevel(as.factor(df$race),
                   ref = "Non-Hispanic White")
```

```{r}
## exploratory data analysis 

df %>%
  group_by(education) %>%
  summarise(prop_cardio = mean(cardio == 1)) %>%
  ggplot(aes(x = education, y = prop_cardio)) +
  geom_col() +
  labs(
    x = "Education Level",
    y = "Proportion with Cardiovascular Disease",
    title = "Cardiovascular Disease by Education Level"
  ) +
  theme_minimal()
```

```{r}
## fitted the model with just the confounders, all have p-values less than 0.2 besides  raceNon-Hispanic Black (p = 0.50062) and raceOther Race - Including Multi-Racial (p = 0.32112), so I'm going to keep all confounder 

model_confounders <- glm(
  cardio ~ age + gender + race + income,
  data = df,
  family = binomial
)
summary(model_confounders)
```

```{r}
model_base <- glm(
  cardio ~ education + age + gender + race + income,
  data = df,
  family = binomial
)
summary(model_base)
```

```{r}
## cardiovascular outcomes based on education really don't change by race 
model_int_race <- glm(
  cardio ~ education * race + age + gender + income,
  data = df,
  family = binomial
)

summary(model_int_race)
```

```{r}
## cardiovascular outcomes based on education do change by gender
model_int_gender <- glm(
  cardio ~ education * gender + age + race + income,
  data = df,
  family = binomial
)

summary(model_int_gender)
```

```{r}
anova(model_base, model_int_gender, test = "LRT")

```

*investigating whether the effect of education differs by gender adjusted for age, race, income*

*education reference = not high school graduate, gender reference = male*

**Among males, there is no statistically significant association between education level and cardiovascular disease**

-   educationCollege Graduate or above 0.0878 p = 0.7017

-   educationHigh School Graduate 0.1549 p = 0.3881

**The education effect for females differs from males by -0.9027 (p = 0.0117 ) (on the log-odds scale)**

-   educationCollege Graduate or above:genderFemale (Estimate = -0.9027, p = 0.0117)

-   female college effect = exp(0.087806 + (-0.9027)) = exp(-0.814894) = 0.4426862

**Among females, college graduates have approximately 56% lower odds of cardiovascular disease compared to women without a high school degree.**

**There is statistical evidence that the association between education and cardiovascular health differs by gender.**
